---
title: "ash-microclimate-data-exploration"
author: "Deidre Jaeger"
date: "10/25/2018"
output: github_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

```


## Exploratory Data Analysis

### new addition: add overlaping histograms to Step 2

ggplot(data=ash_data) +
    geom_histogram(mapping = aes(x=tempF,y=stat(density),fill=ash_removal),
                   position="identity",bins=36,alpha=0.6) 


*Step 1:* load data and make sure it has correct date, number, and factor classes

*Step 2:* view the histograms to see the distribution of temperatures of before and after treatments
 -are there any erroneous measurements? 
 
*Step 3:* time series plot for each of the 4 sensors (x-axis = time, y axis = temperature F)

*Step 4:* combined time series plots for all 3 groups (treatment-tree, control-tree, control-sun), 1 for each treatment

*Further analysis- still in progress*
consider average the two ash treatment replicates?
  
```{r load-libraries, echo=FALSE}
library(dplyr) # for mutate() functionality
library(lubridate) # for as.POSIX function
library(ggplot2)
options(stringsAsFactors = FALSE) # so that dates read in as character strings not factors (factors are troublesome for converting to dates)

```
  
```{r step1-load-data, echo=FALSE}

# read csv file
ash_data <- read.csv("data/ash-microclimate-data2018-prelim.csv", header = T) 

# check structure and make sure every column and row we expect to be there is present
head(ash_data)
unique(ash_data$sensor_name)
str(ash_data) # darn, we dont have the factors and date formatted anymore

# reset date and factors
ash_data <- ash_data %>% 
  mutate(date_time = ymd_hms(date_time), sensor_sn = as.factor(sensor_sn), sensor_name = as.factor(sensor_name), ash_removal = as.factor(ash_removal))

# recheck structure to make sure date=POSIXct, tempF = num, and everything else is a factor- except we don't care about the "X" factor
str(ash_data) 

```

```{r step2-view-histograms-by-sensor}

ash_data %>% # define date
  ggplot() +
  geom_histogram(mapping = aes(x= tempF), bins = 40) + # define x axis
  facet_grid(facets = sensor_name ~ ash_removal) # define which variables to panel vertical and horizontal graphs with

```

```{r step3-view-time-series-scatterplot}

ash_data %>%
  filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% 
  group_by(ash_removal) %>% 
  ggplot() +
  geom_point(mapping = aes(x = date_time, y = tempF,
                           color = ash_removal)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59")) +
  facet_wrap( facets = ~ sensor_name)

```

```{r step3-view-time-series-lineplot}

ash_data %>%
  filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% 
  group_by(ash_removal) %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
                           color = ash_removal)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59")) +
  facet_wrap( facets = ~ sensor_name)

```


```{r step4-overlayed-lineplots}

```

```{r rename-factors-for-plotting}

# create an empty column
ash_data1 <- ash_data %>% 
  mutate(ash_removal_plot = NA)

# fill the new column with the ash removal status that starts with a number so it will order numerically as a factor
ash_data1.1 <- ash_data1 %>% 
  filter(ash_removal == "before_removal") %>% 
  mutate(ash_removal_plot = "1_before_removal")

ash_dat1.2 <- ash_data1 %>% 
  filter(ash_removal == "during_removal") %>% 
  mutate(ash_removal_plot = "2_during_removal")

ash_dat1.3 <- ash_data1 %>% 
  filter(ash_removal == "after_removal") %>% 
  mutate(ash_removal_plot = "3_after_removal")

# fuse the dataframes
ash_data_all <- rbind(ash_data1.1, ash_dat1.2, ash_dat1.3)

```


```{r histograms-by-sensor, echo = FALSE}
ash_data %>% ggplot() + geom_histogram(mapping = aes(x = tempF, color = ash_removal), bins = 40) + facet_grid(facets = sensor_name ~ ash_removal)

ash_data_all %>% 
  ggplot() + 
  geom_histogram(mapping = aes(x = tempF, color = ash_removal), bins = 40) + 
  facet_grid(facets = sensor_name ~ ash_removal_plot)
```

make a histogram with before and after data, separated by sensor (basically a histogram of the last plot)

```{r overlapping-histograms, echo = FALSE}
ggplot(data = ash_data) +
  geom_histogram(mapping = aes(x = tempF, y = stat(density), fill = ash_removal), position = "identity", bins = 36, alpha = 0.6)
```

```{r time-series-scatterplot, echo = FALSE}
ash_data %>% filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% group_by(ash_removal) %>% 
  ggplot() + 
  geom_point(mapping = aes(x = date_time, y = tempF, color = ash_removal)) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 12:20:59")) + 
  facet_wrap(facets = ~ sensor_name)
```

```{r time-series-lineplot, echo = FALSE}
ash_data %>% filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% group_by(ash_removal) %>% 
  ggplot() + 
  geom_line(mapping = aes(x = date_time, y = tempF, color = ash_removal)) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 12:20:59")) + 
  facet_wrap(facets = ~ sensor_name)
```

```{r overlay-line-plots, echo = FALSE}
ash_data %>% filter(ash_removal != "during_removal") %>%
  group_by(ash_removal) %>% 
  ggplot() + 
  geom_line(mapping = aes(x = date_time, y = tempF, color = sensor_name)) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 12:20:59")) 
```

```{r overlay-line-plots-2, echo = FALSE}
ash_data %>% filter(ash_removal != "during_removal") %>% #try filtering out dates with no data
  group_by(ash_removal) %>%
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF, color = sensor_name)) +
  facet_wrap(facets = ~ ash_removal) #check facet_wrap help for how to order the facets
```
