---
title: "ash-microclimate-data-exploration"
author: "Deidre Jaeger"
date: "10/25/2018"
output: github_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

```


## Exploratory Data Analysis

*Step 1:* load data and make sure it has correct date, number, and factor classes

*Step 2:* view the histograms to see the distribution of temperatures of before and after treatments
 -are there any erroneous measurements?
 
*Step 3:* time series plot for each of the 4 sensors (x-axis = time, y axis = temperature F)

*Step 4:* combined time series plots for all 3 groups (treatment-tree, control-tree, control-sun), 1 for each treatment

*Further analysis- still in progress*
consider average the two ash treatment replicates?
  
```{r load-libraries, echo=FALSE}
library(dplyr) # for mutate() functionality
library(lubridate) # for as.POSIX function
library(ggplot2)
options(stringsAsFactors = FALSE) # so that dates read in as character strings not factors (factors are troublesome for converting to dates)

```
  
```{r step1-load-data, echo=FALSE}

# read csv file
ash_data <- read.csv("data/ash-microclimate-data2018.csv", header = T) 

# check structure and make sure every column and row we expect to be there is present
head(ash_data)
unique(ash_data$sensor_name)
str(ash_data) # darn, we dont have the factors and date formatted anymore

# reset date and factors
ash_data <- ash_data %>% 
  mutate(date_time = ymd_hms(date_time), sensor_sn = as.factor(sensor_sn), sensor_name = as.factor(sensor_name), ash_removal = as.factor(ash_removal))

# recheck structure to make sure date=POSIXct, tempF = num, and everything else is a factor- except we don't care about the "X" factor
str(ash_data) 

```

```{r step2-view-histograms-by-sensor}

ash_data %>% # define date
  ggplot() +
  geom_histogram(mapping = aes(x= tempF), bins = 40) + # define x axis
  facet_grid(facets = sensor_name ~ ash_removal) # define which variables to panel vertical and horizontal graphs with

```

```{r step3-view-time-series-scatterplot}

ash_data %>%
  filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% 
  group_by(ash_removal) %>% 
  ggplot() +
  geom_point(mapping = aes(x = date_time, y = tempF,
                           color = ash_removal)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59")) +
  facet_wrap( facets = ~ sensor_name)

```

```{r step4-view-time-series-lineplot}

ash_data %>%
  filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% 
  group_by(ash_removal) %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
                           color = ash_removal)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59")) +
  facet_wrap( facets = ~ sensor_name)

```



