---
title: "02_ash_microclimate_data_analysis"
author: "Amanda Looze"
date: "November 30, 2018"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries-and-data, echo = FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
options(stringsAsFactors = FALSE)

treatment_100e_clean <- read.csv("data/treatment_100e_clean.csv", header = T)
treatment_101w_clean <- read.csv("data/treatment_101w_clean.csv", header = T)
control_sun_data_clean <- read.csv("data/control_sun_data_clean.csv", header = T)
control_sun_data_clean <- control_sun_data_clean %>% filter(ash_removal != "during_removal")
control_103_data_clean <- read.csv("data/control_103_data_clean.csv", header = T)
control_103_data_clean <- control_103_data_clean %>% filter(ash_removal != "during_removal")
```
*Step 1:* Calculate the difference in temp between the treatments and controls for the study duration 
##new dataframe or new columns to current dataframe
```{r calc-diff-temps}
t100e_diff <- data.frame(cbind(treatment_100e_clean$tempF, control_sun_data_clean$tempF, control_103_data_clean$tempF))
colnames(t100e_diff) <- c("t100e", "csun", "c103")

t100e_diff$diff_100e_sun <- t100e_diff$csun - t100e_diff$t100e
t100e_diff$diff_100e_103 <- t100e_diff$c103 - t100e_diff$t100e

##############################################################################

t101w_diff <- data.frame(cbind(treatment_101w_clean$tempF, control_sun_data_clean$tempF, control_103_data_clean$tempF))
colnames(t101w_diff) <- c("t101w", "csun", "c103")

t101w_diff$diff_101w_sun <- t101w_diff$csun - t101w_diff$t101w
t101w_diff$diff_101w_103 <- t101w_diff$c103 - t101w_diff$t101w
```

*Step 2:* Run a paired t-test between the hourly averaged differences in temperature before ash removal and after 

*Step 3:* Make a barplot or boxplot to show the average difference and st error in temperature in the before and after treatments. 
```{r cars}
summary(cars)
```

```{r pressure, echo=FALSE}
plot(pressure)
```

