---
title: "02_ash_microclimate_data_analysis"
author: "Amanda Looze"
date: "November 30, 2018"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries, echo = FALSE}
library(plyr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(reshape2)
options(stringsAsFactors = FALSE)
```

```{r load-data, echo = FALSE}
ash_data <- read.csv("data/ash-microclimate-data2018.csv", header = T)
```


*Step 1:* Calculate the difference in temp between the treatments and controls for the study duration ---DONE
```{r temp-treatment-control-diffs, echo = FALSE}
ash_data2 <- ash_data %>% filter(ash_removal != "during_removal")

unique(ash_data2$ash_removal)


tt_100e_before <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_100e", ash_removal == "before_removal")

tt_101w_before <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_101w", ash_removal == "before_removal")

c_sun_before <- ash_data2 %>% 
  filter(sensor_name == "control_sun", ash_removal == "before_removal")

c_103_before <- ash_data2 %>% 
  filter(sensor_name == "control_tree_103", ash_removal == "before_removal")

ash_data3_before <- tt_100e_before %>% 
  inner_join(tt_101w_before, by = "date_time") %>% 
  inner_join(c_sun_before, by = "date_time") %>% 
  inner_join(c_103_before, by = "date_time") %>% 
  rename(tempF.100e = tempF.x, tempF.101w = tempF.y, tempF.sun = tempF.x.x, tempF.103 = tempF.y.y) %>% 
  select(date_time, ash_removal.x, tempF.100e, tempF.101w, tempF.sun, tempF.103) %>% 
  mutate(diff_sun_100e = (tempF.sun-tempF.100e), 
         diff_sun_101w = (tempF.sun-tempF.101w), 
         diff_sun_103 = (tempF.sun-tempF.103), 
         diff_103_100e = (tempF.103-tempF.100e), 
         diff_103_101w = (tempF.103-tempF.101w), 
         diff_100e_101w = (tempF.100e-tempF.101w))

head(ash_data3_before)

str(ash_data3_before)




tt_100e_after <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_100e", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

tt_101w_after <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_101w", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

c_sun_after <- ash_data2 %>% 
  filter(sensor_name == "control_sun", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

c_103_after <- ash_data2 %>% 
  filter(sensor_name == "control_tree_103", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

ash_data3_after <- tt_100e_after %>% 
  inner_join(tt_101w_after, by = "date_time2") %>% 
  inner_join(c_sun_after, by = "date_time2") %>% 
  inner_join(c_103_after, by = "date_time2") %>% 
  rename(tempF.100e = tempF.x, tempF.101w = tempF.y, tempF.sun = tempF.x.x, tempF.103 = tempF.y.y) %>%
  select(date_time2, ash_removal.x, tempF.100e, tempF.101w, tempF.sun, tempF.103) %>% 
  mutate(diff_sun_100e = (tempF.sun-tempF.100e), 
         diff_sun_101w = (tempF.sun-tempF.101w), 
         diff_sun_103 = (tempF.sun-tempF.103), 
         diff_103_100e = (tempF.103-tempF.100e), 
         diff_103_101w = (tempF.103-tempF.101w), 
         diff_100e_101w = (tempF.100e-tempF.101w))

head(ash_data3_after)

str(ash_data3_after)
```


*Step 2.1:*  1) calculate some summary statistics
            - for each treatment pair
               a) what is the max temp difference, at what time of the day
               b) what is the min temp diff, at what time of day?
               c) save to a table #as.table()
```{r summary-stats, echo=FALSE}
maxb_sun_100e <- max(ash_data3_before[1:4325,7])
p_sun_100e <- which(ash_data3_before$diff_sun_100e == maxb_sun_100e)
dtb_sun_100e <- ash_data3_before[p_sun_100e,1]

#####################

maxb_sun_101w <- max(ash_data3_before[1:4325,8])
p_sun_101w <- which(ash_data3_before$diff_sun_101w == maxb_sun_101w)
dtb_sun_101w <- ash_data3_before[p_sun_101w,1]

# max_values <- cbind(dtb_sun_100e, maxb_sun_100e,dtb_sun_101w, maxb_sun_101w)

#####################

maxb_sun_103 <- max(ash_data3_before[1:4325,9])
p_sun_103 <- which(ash_data3_before$diff_sun_103 == maxb_sun_103)
dtb_sun_103 <- ash_data3_before[p_sun_103,1]

#####################

maxb_103_100e <- max(ash_data3_before[1:4325,10])
p_103_100e <- which(ash_data3_before$diff_103_100e == maxb_103_100e)
dtb_103_100e <- ash_data3_before[p_103_100e,1]

#####################

maxb_103_101w <- max(ash_data3_before[1:4325,11])
p_103_101w <- which(ash_data3_before$diff_103_101w == maxb_103_101w)
dtb_103_101w <- ash_data3_before[p_103_101w,1]

#####################

maxb_100e_101w <- max(ash_data3_before[1:4325,12])
p_100e_101w <- which(ash_data3_before$diff_100e_101w == maxb_100e_101w)
dtb_100e_101w <- ash_data3_before[p_100e_101w,1]

max_values <- cbind(dtb_sun_100e, maxb_sun_100e, dtb_sun_101w, maxb_sun_101w, dtb_sun_103, maxb_sun_103, dtb_103_100e, maxb_103_100e, dtb_103_101w, maxb_103_101w, dtb_100e_101w, maxb_100e_101w)

# extreme_data <- matrix(c(), ncol = 5)
# colnames(extreme_data) <- c("date_time", "before_after", "max_min", "sensors", "value")
# extreme_data <- as.table(extreme_data)
# str(extreme_data)

####################################################### for loop

# for (year in c(2010:2015)){print(paste("the year is", year))}
# tester <- apply(d1, FUN=which.max, 2)
# for(maxbefore in c(ash_data3_before[1:4325, 7:12])
#     {print(paste("the max is", maxbefore))}

# test <- apply(ash_data3_before, FUN = which.max, 7)
```

*Step 2.1.1:*
2) show some summary graphs 
   a) plot histograms of the difference values between treatments (using ggplot facet)---DONE
```{r overlapping-histograms, echo = FALSE}

ash_data_diffs <- data.frame("diff_cat" = rep(NA, 51900), "diff_tempF" = rep(NA, 51900), "rem_stat" = rep(NA, 51900))

ash_data_diffs$diff_cat <- c(rep("sun_100e", 4325), rep("sun_101w", 4325), rep("sun_103", 4325), rep("103_100e", 4325), rep("103_101w", 4325), rep("100e_101w", 4325), rep("sun_100e", 4325), rep("sun_101w", 4325), rep("sun_103", 4325), rep("103_100e", 4325), rep("103_101w", 4325), rep("100e_101w", 4325))

ash_data_diffs$diff_cat <- as.factor(ash_data_diffs$diff_cat)

ash_data_diffs$rem_stat <- c(rep("before_removal", 25950), rep("after_removal", 25950))

ash_data_diffs$diff_tempF[1:4325] <- c(ash_data3_before[,7])
ash_data_diffs$diff_tempF[4326:8650] <- c(ash_data3_before[,8])
ash_data_diffs$diff_tempF[8651:12975] <- c(ash_data3_before[,9])
ash_data_diffs$diff_tempF[12976:17300] <- c(ash_data3_before[,10])
ash_data_diffs$diff_tempF[17301:21625] <- c(ash_data3_before[,11])
ash_data_diffs$diff_tempF[21626:25950] <- c(ash_data3_before[,12])

ash_data_diffs$diff_tempF[25951:30275] <- c(ash_data3_after[,7])
ash_data_diffs$diff_tempF[30276:34600] <- c(ash_data3_after[,8])
ash_data_diffs$diff_tempF[34601:38925] <- c(ash_data3_after[,9])
ash_data_diffs$diff_tempF[38926:43250] <- c(ash_data3_after[,10])
ash_data_diffs$diff_tempF[43251:47575] <- c(ash_data3_after[,11])
ash_data_diffs$diff_tempF[47576:51900] <- c(ash_data3_after[,12])

ash_data_diffs$rem_stat <- factor(ash_data_diffs$rem_stat, levels = c('before_removal','after_removal'))

ggplot(data = ash_data_diffs) +
  geom_density(mapping = aes(x = diff_tempF, y = stat(density), color = diff_cat), position = "identity", bins = 36, alpha = 0.6) +
  facet_wrap(facets = ~ rem_stat) 

ggplot(data = ash_data_diffs) +
  geom_density(mapping = aes(x = diff_tempF, y = stat(density), color = diff_cat), position = "identity", bins = 36, alpha = 0.6) +
  xlim(-3, 3) +
  facet_wrap(facets = ~ rem_stat) 
```
               

*Step 2.2:* Run overall value t-tests between the hourly averaged differences in temperature before ash removal and after ---DONE
```{r t-test, echo = FALSE}
ash_data3_before[4326:5497, ] <- NA

t_sun_100e <- t.test(ash_data3_after$diff_sun_100e, 
                    ash_data3_before$diff_sun_100e, paired = TRUE)

psun100e <- t_sun_100e$p.value

paste("p-val for sun vs 100e is ", psun100e)

#########

t_sun_101w <- t.test(ash_data3_after$diff_sun_101w, 
                    ash_data3_before$diff_sun_101w, paired = TRUE)

psun101w <- t_sun_101w$p.value

paste("p-val for sun vs 101w is ", psun101w)

#########

t_sun_103 <- t.test(ash_data3_after$diff_sun_103, 
                    ash_data3_before$diff_sun_103, paired = TRUE)

psun103 <- t_sun_103$p.value

paste("p-val for sun vs 103 is ", psun103)

#########

t_103_100e <- t.test(ash_data3_after$diff_103_100e, 
                    ash_data3_before$diff_103_100e, paired = TRUE)

p103100e <- t_103_100e$p.value

paste("p-val for 103 vs 100e is ", p103100e)

#########

t_103_101w <- t.test(ash_data3_after$diff_103_101w, 
                    ash_data3_before$diff_103_101w, paired = TRUE)

p103101w <- t_103_101w$p.value

paste("p-val for 103 vs 101w is ", p103101w)

#########

t_100e_101w <- t.test(ash_data3_after$diff_100e_101w, 
                    ash_data3_before$diff_100e_101w, paired = TRUE)

p100e101w <- t_100e_101w$p.value

paste("p-val for 100e vs 101w is ", p100e101w)
```


*Step 3:* Make a barplot or boxplot to show the average difference and st error in mean temperatures in the before and after treatments.---DONE
```{r bar-plot-avg-diffs, echo=FALSE}

melted <- melt(ash_data_diffs, id.vars = c("diff_cat", "rem_stat"))

mean_diffs <- ddply(melted, c("diff_cat", "rem_stat"), summarise, mean = mean(value), se = sd(value))

mean_diffs$diff_cat <- as.factor(mean_diffs$diff_cat)

ggplot(data = mean_diffs, aes(x = diff_cat, y = mean, color = diff_cat, fill = diff_cat)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width = .2) +
  facet_wrap(facets = ~rem_stat)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```


*Step 4:* Make a line plot of the differences in temperature in the before and after treatments #in relation to days
    - may need to change "date_time" to day 1, 2, 3 so able to fully overlay values
```{r line-plot-diffs, echo = FALSE}
ash_data3_after2 <- ash_data3_after %>% 
  rename(date_time = date_time2)

 

#c("date_time", "ash_removal.x", "tempF.100e", "tempF.101w", "tempF.sun", "tempF.103", "diff_sun_100e", "diff_sun_101w", "diff_sun_103", "diff_103_100e", "diff_103_101w", "diff_100e_101w")


ash_data3 <- rbind(ash_data3_before, ash_data3_after2)


ash_data3 %>%
  group_by(ash_removal) %>%
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF, color = sensor_name)) +
  facet_wrap(facets = ~ ash_removal)
```

