---
title: "02_ash-microclimate-data-analysis"
author: "Deidre Jaeger"
date: "11/29/2018"
output: pdf_document
---

## Analysis Plan

*Step 1:* Calculate the difference in temp between the treatments and controls for the study duration

*Step 2:* 1) calculate some summary statistics
            - for each treatment pair
               a) what is the max temp difference, at what time of the day
               b) what is the min temp diff, at what time of day?
               c) save to a table
              
          2) show some summary graphs
              a) plot histograms of the difference values between treatments (using ggplot facet)
  

*Step 2:* Run a overall value t-tests between the hourly averaged differences in temperature before ash removal and after 

*Step 3:* Make a barplot or boxplot to show the average difference and st error in mean temperatures in the before and after treatments. 

*Step 4:* Make a line plot of the differences in temperature in the before and after treatments 
    - may need to change "date_time" to day 1, 2, 3 so able to fully overlay values



```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries-and-data, echo = FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
options(stringsAsFactors = FALSE)

```


```{r load-data}
# read csv file
ash_data <- read.csv("data/ash-microclimate-data2018.csv", header = T) 
```


```{r step-1-temp-treatment-control-diffs}
# create a copy of the dataset and prep for subtractions
ash_data2 <- ash_data %>%
  filter(ash_removal != "during_removal") # remove any results "during removal"

# verify that "during" values were removed from ash_data2
unique(ash_data2$ash_removal)

# # spread data into 4 columns for temp values [not working!! Maybe due to the times being off at some point during the study?]
# ash_data2 %>% 
#   filter(sensor_name == "control_sun" | sensor_name == "treatment_tree_100e") %>% 
#   spread(key = sensor_name, value = tempF)

# create new temp dataframes for values
tt_100e_before <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_100e", ash_removal == "before_removal")

tt_101w_before <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_101w", ash_removal == "before_removal")

c_sun_before <- ash_data2 %>% 
  filter(sensor_name == "control_sun", ash_removal == "before_removal")

c_103_before <- ash_data2 %>% 
  filter(sensor_name == "control_tree_103", ash_removal == "before_removal")


# get all the "before" sensor data into separate columns and subtract temp differences
ash_data3_before <- tt_100e_before %>%  # name a new dataframe
  inner_join(tt_101w_before, by = "date_time") %>% #join 101w to 100e by date_time
  inner_join(c_sun_before, by = "date_time") %>% #join c_sun to 100e by date_time
  inner_join(c_103_before, by = "date_time") %>% #join c_103 to 100e by date_time
  rename(tempF.100e = tempF.x, tempF.101w = tempF.y, tempF.sun = tempF.x.x, tempF.103 = tempF.y.y) %>% # rename columns
  select(date_time, ash_removal.x, tempF.100e, tempF.101w, tempF.sun, tempF.103) %>% # pull out the columns that are important
  mutate(diff_sun_100e = (tempF.sun-tempF.100e), 
         diff_sun_101w = (tempF.sun-tempF.101w), 
         diff_sun_103 = (tempF.sun-tempF.103), 
         diff_103_100e = (tempF.103-tempF.100e), 
         diff_103_101w = (tempF.103-tempF.101w), 
         diff_100e_101w = (tempF.100e-tempF.101w))   # create new columns to store the difference values


# the after removal times aren't matching up so need to cut off the second decimal places
tt_100e_after <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_100e", ash_removal == "after_removal") %>% 
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

tt_101w_after <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_101w", ash_removal == "after_removal") %>% 
    mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))
  

c_sun_after <- ash_data2 %>% 
  filter(sensor_name == "control_sun", ash_removal == "after_removal") %>% 
    mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))


c_103_after <- ash_data2 %>% 
  filter(sensor_name == "control_tree_103", ash_removal == "after_removal") %>% 
    mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))


# get all the "after" sensor data into separate columns and subtract temp differences
ash_data3_after <- tt_100e_after %>%  # name a new dataframe
  inner_join(tt_101w_after, by = "date_time2") %>% #join 101w to 100e by date_time
  inner_join(c_sun_after, by = "date_time2") %>% #join c_sun to 100e by date_time
  inner_join(c_103_after, by = "date_time2") %>% #join c_103 to 100e by date_time
  rename(tempF.100e = tempF.x, tempF.101w = tempF.y, tempF.sun = tempF.x.x, tempF.103 = tempF.y.y) %>% # rename columns
  select(date_time2, ash_removal.x, tempF.100e, tempF.101w, tempF.sun, tempF.103) %>% # pull out the columns that are important
  mutate(diff_sun_100e = (tempF.sun-tempF.100e), diff_sun_101w = (tempF.sun-tempF.101w), diff_sun_103 = (tempF.sun-tempF.103), diff_103_100e = (tempF.103-tempF.100e), diff_103_101w = (tempF.103-tempF.101w), diff_100e_101w = (tempF.100e-tempF.101w))   # create new columns to store the difference values



```

*Step 2.1:*  1) calculate some summary statistics
            - for each treatment pair
               a) what is the max temp difference, at what time of the day
               b) what is the min temp diff, at what time of day?
               c) save to a dataframe
              
          ###2) show some summary graphs #
              a) plot histograms of the difference values between treatments (using ggplot facet)
```{r step2-summary-stats-before-removal}
# what is the max temp difference between sun and 100e treatments
maxb_sun_100e <- abs(max(ash_data3_before[1:4325,7])) # max diff is 10.19 degrees F
# at what time of the day does this difference occur?
p_sun_100e <- which(ash_data3_before$diff_sun_100e == maxb_sun_100e)
dtb_sun_100e <- ash_data3_before[p_sun_100e,1] # on Sept 17, 8:49 AM
# what is the min temp difference between sun and 100e treatments
minb_sun_100e <- abs(min(ash_data3_before[1:4325,7])) # max diff is 10.19 degrees F
# at what time of the day does this difference occur?
p2_sun_100e <- which(ash_data3_before$diff_sun_100e == minb_sun_100e)
dtb2_sun_100e <- ash_data3_before[p2_sun_100e,1] # on Sept 17, 8:49 AM

#####################
# what is the max temp difference between sun and 100w treatments
maxb_sun_101w <- abs(max(ash_data3_before[1:4325,8])) # max diff is 8.11 degrees F
# at what time of the day does this difference occur?
p_sun_101w <- which(ash_data3_before$diff_sun_101w == maxb_sun_101w)
dtb_sun_101w <- ash_data3_before[p_sun_101w,1]

# summary of values
max_values <- cbind(dtb_sun_100e, maxb_sun_100e,dtb_sun_101w, maxb_sun_101w)

#####################
# what is the max temp difference between sun and 103 treatments?
maxb_sun_103 <- abs(max(ash_data3_before[1:4325,9]))
p_sun_103 <- which(ash_data3_before$diff_sun_103 == maxb_sun_103)
dtb_sun_103 <- ash_data3_before[p_sun_103,1]

#####################
# what is the max temp difference between 103 and 100e treatments?
maxb_103_100e <- abs(max(ash_data3_before[1:4325,10]))
p_103_100e <- which(ash_data3_before$diff_103_100e == maxb_103_100e)
dtb_103_100e <- ash_data3_before[p_103_100e,1]

#####################
# what is the max temp difference between 103 and 101w treatments?
maxb_103_101w <- abs(max(ash_data3_before[1:4325,11]))
p_103_101w <- which(ash_data3_before$diff_103_101w == maxb_103_101w)
dtb_103_101w <- ash_data3_before[p_103_101w,1]

#####################
# what is the max temp difference between 100e and 101w treatments?
maxb_100e_101w <- abs(max(ash_data3_before[1:4325,12]))
p_100e_101w <- which(ash_data3_before$diff_100e_101w == maxb_100e_101w)
dtb_100e_101w <- ash_data3_before[p_100e_101w,1]


# max_v <- as.data.frame(diff_between = c("dtb_sun_100e", "dtb_sun_101w", "dtb_sun_103", "dtb_103_100e", "dtb_103_101w", "dtb_100e_101w"), max_temp_diff = c(maxb_sun_100e, maxb_sun_101w, maxb_sun_103, maxb_103_100e, maxb_103_101w, maxb_100e_101w), time = c(dtb_sun_100e, dtb_sun_101w, dtb_sun_103, dtb_103_100e, dtb_103_101w, dtb_100e_101w))

##################################################################
# create a dataframe with the difference mins and maxs, and the time they occur

diff_between = c("sun_100e", "sun_101w", "sun_103", "103_100e", "103_101w", "100e_101w")

max_v1 <- as.data.frame(diff_between)
max_v1$max_temp_diff_b4R <- c(maxb_sun_100e, maxb_sun_101w, maxb_sun_103, maxb_103_100e, maxb_103_101w, maxb_100e_101w)
max_v1$time_b4R <- c(dtb_sun_100e, dtb_sun_101w, dtb_sun_103, dtb_103_100e, dtb_103_101w[[1]], dtb_100e_101w[[1]])

```

```{r step2-summary-stats-after-removal}
 ####################################################### 

# calc for after removal

# what is the max temp difference between sun and 100e treatments
maxa_sun_100e <- abs(max(ash_data3_after[,7]))
# at what time of the day does this difference occur?
r_sun_100e <- which(ash_data3_after$diff_sun_100e == maxa_sun_100e)
dta_sun_100e <- ash_data3_after[r_sun_100e,1] 

#####################
# what is the max temp difference between sun and 100w treatments
maxa_sun_101w <- abs(max(ash_data3_after[,8])) 
# at what time of the day does this difference occur?
r_sun_101w <- which(ash_data3_after$diff_sun_101w == maxa_sun_101w)
dta_sun_101w <- ash_data3_after[r_sun_101w,1]

#####################
# what is the max temp difference between sun and 103 treatments?
maxa_sun_103 <- abs(max(ash_data3_after[,9]))
r_sun_103 <- which(ash_data3_after$diff_sun_103 == maxa_sun_103)
dta_sun_103 <- ash_data3_after[r_sun_103,1]

#####################
# what is the max temp difference between 103 and 100e treatments?
maxa_103_100e <- abs(max(ash_data3_after[,10]))
r_103_100e <- which(ash_data3_after$diff_103_100e == maxa_103_100e)
dta_103_100e <- ash_data3_after[r_103_100e,1]

#####################
# what is the max temp difference between 103 and 101w treatments?
maxa_103_101w <- abs(max(ash_data3_after[,11]))
r_103_101w <- which(ash_data3_after$diff_103_101w == maxa_103_101w)
dta_103_101w <- ash_data3_after[r_103_101w,1]

#####################
# what is the max temp difference between 100e and 101w treatments?
maxa_100e_101w <- abs(max(ash_data3_after[,12]))
r_100e_101w <- which(ash_data3_after$diff_100e_101w == maxa_100e_101w)
dta_100e_101w <- ash_data3_after[r_100e_101w,1]


##################################################################
# create a dataframe with the difference mins and maxs, and the time they occur
# 
# diff_aft_removal = c("dta_sun_100e", "dta_sun_101w", "dta_sun_103", "dta_103_100e", "dta_103_101w", "dta_103_101w", "dta_100e_101w", "dta_100e_101w")

max_v1$max_temp_diff_aftR <- c(maxa_sun_100e, maxa_sun_101w, maxa_sun_103, maxa_103_100e, maxa_103_101w, maxa_100e_101w)

max_v1$time_aftR <- c(dta_sun_100e, dta_sun_101w, dta_sun_103, dta_103_100e, dta_103_101w[[1]], dta_100e_101w)


####################################################### for loop

```
Consider looking at the min temp diffs and the time, write a function to calculate it? 

```{r step-3-t-tests}
# overall t-tests using full columns of data
t.test(ash_data3_before$tempF.sun, ash_data3_before$tempF.100e) # temps between sun control and 100e treatment signficantly different
t.test(ash_data3_after$tempF.sun, ash_data3_after$tempF.100e) # not significantly different

t.test(ash_data3_before$diff_sun_100e, ash_data3_after$diff_sun_100e) # the mean difference in "before removal" temperature between sun control and 100e is signficantly different from the "after removals" temperature between sun control and 100e

# write out to a table for comparison


# not sure if we need paried tests
# # use paired t-tests to compare the difference values on a minute by minute basis
# t.test(ash_data3_before$tempF.sun, ash_data3_before$tempF.100e, paired = TRUE, conf.level = 0.95)
# t.test(ash_data3_after$tempF.sun, ash_data3_after$tempF.100e, paired = TRUE, conf.level = 0.95)


```


*Step 2.2:* Run overall value t-tests between the hourly averaged differences in temperature before ash removal and after
```{r t-test, echo = FALSE}
ash_data3_before[4326:5497, ] <- NA

t_sun_100e <- t.test(ash_data3_after$diff_sun_100e, 
                    ash_data3_before$diff_sun_100e, paired = TRUE)

t_sun_100e$p.value

#########

t_sun_101w <- t.test(ash_data3_after$diff_sun_101w, 
                    ash_data3_before$diff_sun_101w, paired = TRUE)

t_sun_101w$p.value

#########

t_sun_103 <- t.test(ash_data3_after$diff_sun_103, 
                    ash_data3_before$diff_sun_103, paired = TRUE)

t_sun_103$p.value

#########

t_103_100e <- t.test(ash_data3_after$diff_103_100e, 
                    ash_data3_before$diff_103_100e, paired = TRUE)

t_103_100e$p.value

#########

t_103_101w <- t.test(ash_data3_after$diff_103_101w, 
                    ash_data3_before$diff_103_101w, paired = TRUE)

t_103_101w$p.value

#########

t_100e_101w <- t.test(ash_data3_after$diff_100e_101w, 
                    ash_data3_before$diff_100e_101w, paired = TRUE)

t_100e_101w$p.value
```

*Step 3:* Make a barplot or boxplot to show the average difference and st error in mean temperatures in the before and after treatments. 
```{r bar-plot-avg-diffs, echo=FALSE}

ash_data3_before %>% filter(ash_data3_before$date_time >= "2018-09-17 00:00:59" | ash_data3_before$date_time < "2018-09-17 01:00:59") %>% summarize(avg = mean(ash_data3_before))

plot(avg)


# ggplot(data = stom_spec, mapping = aes(x = species, y = stomata_density)) +
#   geom_bar(stat = "identity", color = area.color, fill = area.color) + 
#   geom_errorbar(aes(ymin = SE_lower, ymax = SE_upper), width = 0.4) + 
#   geom_text(aes(x = species, y = SE_upper+.4, label = letters), size = 5)+
#   ylab("Stomata Density (1/mm^2)") +
#   xlab("Species") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*Step 4:* ###Make a line plot of the differences in temperature in the before and after treatments 
    - may need to change "date_time" to day 1, 2, 3 so able to fully overlay values
```{r line-plot-diffs, echo = FALSE}
ash_data3_after2 <- ash_data3_after %>% 
  rename(date_time = date_time2)

 

#c("date_time", "ash_removal.x", "tempF.100e", "tempF.101w", "tempF.sun", "tempF.103", "diff_sun_100e", "diff_sun_101w", "diff_sun_103", "diff_103_100e", "diff_103_101w", "diff_100e_101w")


ash_data3 <- rbind(ash_data3_before, ash_data3_after2)

# for this plot, we need to use dplyr to gather the temp values into a single column for the plot

ash_data3 %>%
  group_by(ash_removal.x) %>%
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF, color = sensor_name)) +
  facet_wrap(facets = ~ ash_removal.x)
```

Left off Jan 16: calc mean temps, look at t .tests, finish plots

