---
title: "02_ash_microclimate_data_analysis"
author: "Amanda Looze"
date: "November 30, 2018"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries-and-data, echo = FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
options(stringsAsFactors = FALSE)

ash_data <- read.csv("data/ash-microclimate-data2018.csv", header = T)
head(ash_data)
unique(ash_data$sensor_name)
str(ash_data)

ash_data <- ash_data %>% mutate(date_time = ymd_hms(date_time), sensor_sn = as.factor(sensor_sn), sensor_name = as.factor(sensor_name), ash_removal = as.factor(ash_removal))

str(ash_data)
```
*Step 1:* Calculate the difference in temp between the treatments and controls for the study duration #new dataframe or new columns to current dataframe
```{r calc-diff-temps}
ash_data %>% filter()

```

*Step 2:* Run a paired t-test between the hourly averaged differences in temperature before ash removal and after 

*Step 3:* Make a barplot or boxplot to show the average difference and st error in temperature in the before and after treatments. 
```{r cars}
summary(cars)
```

```{r pressure, echo=FALSE}
plot(pressure)
```

