---
title: "02_ash_dataanalysis_alooze"
author: "Deidre Jaeger"
date: "2/28/2019"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries, echo = FALSE}
library(plyr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(reshape2)
options(stringsAsFactors = FALSE)
```

```{r load-data, echo = FALSE}
ash_data <- read.csv("data/ash-microclimate-data2018-prelim.csv", header = T)
```


*Step 1:* Calculate the difference in temp between the treatments and controls for the study duration 
```{r temp-treatment-control-diffs, echo = FALSE}
unique(ash_data$sensor_name)

ash_data2 <- ash_data %>% filter(ash_removal != "during_removal")

unique(ash_data2$ash_removal)
unique(ash_data2$sensor_name)

tt_100e_before <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_100e", ash_removal == "before_removal")

tt_101w_before <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_101w", ash_removal == "before_removal")

c_sun_before <- ash_data2 %>% 
  filter(sensor_name == "control_sun", ash_removal == "before_removal")

c_103_before <- ash_data2 %>% 
  filter(sensor_name == "control_tree_103", ash_removal == "before_removal")

ash_data3_before <- tt_100e_before %>% 
  inner_join(tt_101w_before, by = "date_time") %>% 
  inner_join(c_sun_before, by = "date_time") %>% 
  inner_join(c_103_before, by = "date_time") %>% #dplyr::rename()
  dplyr::rename(tempF.100e = tempF.x, tempF.101w = tempF.y, tempF.sun = tempF.x.x, tempF.103 = tempF.y.y) %>% 
  select(date_time, ash_removal.x, tempF.100e, tempF.101w, tempF.sun, tempF.103) %>% 
  mutate(diff_sun_100e = (tempF.sun-tempF.100e), 
         diff_sun_101w = (tempF.sun-tempF.101w), 
         diff_sun_103 = (tempF.sun-tempF.103), 
         diff_103_100e = (tempF.103-tempF.100e), 
         diff_103_101w = (tempF.103-tempF.101w), 
         diff_100e_101w = (tempF.100e-tempF.101w))

head(ash_data3_before)

str(ash_data3_before)




tt_100e_after <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_100e", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

tt_101w_after <- ash_data2 %>% 
  filter(sensor_name == "treatment_tree_101w", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

c_sun_after <- ash_data2 %>% 
  filter(sensor_name == "control_sun", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

c_103_after <- ash_data2 %>% 
  filter(sensor_name == "control_tree_103", ash_removal == "after_removal") %>%
  mutate(date_time2 = as.POSIXct(date_time, format = "%Y-%m-%d %H:%M"))

ash_data3_after <- tt_100e_after %>% 
  inner_join(tt_101w_after, by = "date_time2") %>% 
  inner_join(c_sun_after, by = "date_time2") %>% 
  inner_join(c_103_after, by = "date_time2") %>% 
  dplyr::rename(tempF.100e = tempF.x, tempF.101w = tempF.y, tempF.sun = tempF.x.x, tempF.103 = tempF.y.y) %>%
  select(date_time2, ash_removal.x, tempF.100e, tempF.101w, tempF.sun, tempF.103) %>% 
  mutate(diff_sun_100e = (tempF.sun-tempF.100e), 
         diff_sun_101w = (tempF.sun-tempF.101w), 
         diff_sun_103 = (tempF.sun-tempF.103), 
         diff_103_100e = (tempF.103-tempF.100e), 
         diff_103_101w = (tempF.103-tempF.101w), 
         diff_100e_101w = (tempF.100e-tempF.101w))

head(ash_data3_after)

str(ash_data3_after)
```


*Step 2.1:*  1) calculate some summary statistics
            - for each treatment pair
               a) what is the max temp difference, at what time of the day
               b) what is the min temp diff, at what time of day?
               c) save to a table #as.table()
```{r summary-stats, echo=FALSE}
maxb_sun_100e <- max(abs(ash_data3_before[1:4325,7]))
p_sun_100e <- which(ash_data3_before$diff_sun_100e == maxb_sun_100e)
dtb_sun_100e <- ash_data3_before[p_sun_100e,1]

#####################

maxb_sun_101w <- max(abs(ash_data3_before[1:4325,8]))
p_sun_101w <- which(ash_data3_before$diff_sun_101w == maxb_sun_101w)
dtb_sun_101w <- ash_data3_before[p_sun_101w,1]

#####################

maxb_sun_103 <- max(abs(ash_data3_before[1:4325,9]))
p_sun_103 <- which(ash_data3_before$diff_sun_103 == maxb_sun_103)
dtb_sun_103 <- ash_data3_before[p_sun_103,1]

#####################

maxb_103_100e <- max(abs(ash_data3_before[1:4325,10]))
p_103_100e <- which(ash_data3_before$diff_103_100e == -maxb_103_100e)
dtb_103_100e <- ash_data3_before[p_103_100e,1]

#####################

maxb_103_101w <- max(abs(ash_data3_before[1:4325,11]))
p_103_101w <- which(ash_data3_before$diff_103_101w == -maxb_103_101w) # negative difference
dtb_103_101w <- ash_data3_before[p_103_101w,1]

#####################

maxb_100e_101w <- max(abs(ash_data3_before[1:4325,12]))
p_100e_101w <- which(ash_data3_before$diff_100e_101w == -maxb_100e_101w)
dtb_100e_101w <- ash_data3_before[p_100e_101w,1]

max_values_b <- cbind(dtb_sun_100e, maxb_sun_100e, dtb_sun_101w, maxb_sun_101w, dtb_sun_103, maxb_sun_103, dtb_103_100e, maxb_103_100e, dtb_103_101w, maxb_103_101w, dtb_100e_101w, maxb_100e_101w)

#######################################################

minb_sun_100e <- min(abs(ash_data3_before[1:4325,7]))
p2_sun_100e <- which(ash_data3_before$diff_sun_100e == minb_sun_100e)
dtb2_sun_100e <- ash_data3_before[p2_sun_100e,1]

#####################

minb_sun_101w <- min(abs(ash_data3_before[1:4325,8]))
p2_sun_101w <- which(ash_data3_before$diff_sun_101w == minb_sun_101w)
dtb2_sun_101w <- ash_data3_before[p2_sun_101w,1]

#####################

minb_sun_103 <- min(abs(ash_data3_before[1:4325,9]))
p2_sun_103 <- which(ash_data3_before$diff_sun_103 == minb_sun_103)
dtb2_sun_103 <- ash_data3_before[p2_sun_103,1]

#####################

minb_103_100e <- min(abs(ash_data3_before[1:4325,10]))
p2_103_100e <- which(ash_data3_before$diff_103_100e == minb_103_100e)
dtb2_103_100e <- ash_data3_before[p2_103_100e,1]

#####################

minb_103_101w <- min(abs(ash_data3_before[1:4325,11]))
p2_103_101w <- which(ash_data3_before$diff_103_101w == minb_103_101w)
dtb2_103_101w <- ash_data3_before[p2_103_101w,1]

#####################

minb_100e_101w <- min(abs(ash_data3_before[1:4325,12]))
p2_100e_101w <- which(ash_data3_before$diff_100e_101w == minb_100e_101w)
dtb2_100e_101w <- ash_data3_before[p2_100e_101w,1]

min_values_b <- cbind(dtb2_sun_100e, minb_sun_100e, dtb2_sun_101w, minb_sun_101w, dtb2_sun_103, minb_sun_103, dtb2_103_100e, minb_103_100e, dtb2_103_101w, minb_103_101w, dtb2_100e_101w, minb_100e_101w)

##############################################################################################################

maxa_sun_100e <- max(abs(ash_data3_after[1:4325,7]))
p3_sun_100e <- which(ash_data3_after$diff_sun_100e == maxa_sun_100e)
dtb3_sun_100e <- ash_data3_after[p3_sun_100e,1]

#####################

maxa_sun_101w <- max(abs(ash_data3_after[1:4325,8]))
p3_sun_101w <- which(ash_data3_after$diff_sun_101w == maxa_sun_101w)
dtb3_sun_101w <- ash_data3_after[p3_sun_101w,1]

#####################

maxa_sun_103 <- max(abs(ash_data3_after[1:4325,9]))
p3_sun_103 <- which(ash_data3_after$diff_sun_103 == maxa_sun_103)
dtb3_sun_103 <- ash_data3_after[p3_sun_103,1]

#####################

maxa_103_100e <- max(abs(ash_data3_after[1:4325,10]))
p3_103_100e <- which(ash_data3_after$diff_103_100e == -maxa_103_100e)
dtb3_103_100e <- ash_data3_after[p3_103_100e,1]

#####################

maxa_103_101w <- max(abs(ash_data3_after[1:4325,11]))
p3_103_101w <- which(ash_data3_after$diff_103_101w == -maxa_103_101w)
dtb3_103_101w <- ash_data3_after[p3_103_101w,1]

#####################

maxa_100e_101w <- max(abs(ash_data3_after[1:4325,12]))
p3_100e_101w <- which(ash_data3_after$diff_100e_101w == -maxa_100e_101w)
dtb3_100e_101w <- ash_data3_after[p3_100e_101w,1]

max_values_a <- cbind(dtb3_sun_100e, maxa_sun_100e, dtb3_sun_101w, maxa_sun_101w, dtb3_sun_103, maxa_sun_103, dtb3_103_100e, maxa_103_100e, dtb3_103_101w, maxa_103_101w, dtb3_100e_101w, maxa_100e_101w)

#######################################################

mina_sun_100e <- min(abs(ash_data3_after[1:4325,7]))
p4_sun_100e <- which(ash_data3_after$diff_sun_100e == mina_sun_100e)
dtb4_sun_100e <- ash_data3_after[p4_sun_100e,1]

#####################

mina_sun_101w <- min(abs(ash_data3_after[1:4325,8]))
p4_sun_101w <- which(ash_data3_after$diff_sun_101w == mina_sun_101w)
dtb4_sun_101w <- ash_data3_after[p4_sun_101w,1]

#####################

mina_sun_103 <- min(abs(ash_data3_after[1:4325,9]))
p4_sun_103 <- which(ash_data3_after$diff_sun_103 == mina_sun_103)
dtb4_sun_103 <- ash_data3_after[p4_sun_103,1]

#####################

mina_103_100e <- min(abs(ash_data3_after[1:4325,10]))
p4_103_100e <- which(ash_data3_after$diff_103_100e == mina_103_100e)
dtb4_103_100e <- ash_data3_after[p4_103_100e,1]

#####################

mina_103_101w <- min(abs(ash_data3_after[1:4325,11]))
p4_103_101w <- which(ash_data3_after$diff_103_101w == mina_103_101w)
dtb4_103_101w <- ash_data3_after[p4_103_101w,1]

#####################

mina_100e_101w <- min(abs(ash_data3_after[1:4325,12]))
p4_100e_101w <- which(ash_data3_after$diff_100e_101w == mina_100e_101w)
dtb4_100e_101w <- ash_data3_after[p4_100e_101w,1]

min_values_a <- cbind(dtb4_sun_100e, mina_sun_100e, dtb4_sun_101w, mina_sun_101w, dtb4_sun_103, mina_sun_103, dtb4_103_100e, mina_103_100e, dtb4_103_101w, mina_103_101w, dtb4_100e_101w, mina_100e_101w)

```



*Step 2.1.1:*
2) show some summary graphs 
   a) plot histograms of the difference values between treatments (using ggplot facet)

```{r overlapping-histograms, echo = FALSE}

ash_data_diffs <- data.frame("date_time" = rep(NA, 58932), "diff_cat" = rep(NA, 58932), "diff_tempF" = rep(NA, 58932), "rem_stat" = rep(NA, 58932))

ash_data_diffs$diff_cat <- c(rep("sun_100e", 4325), rep("sun_101w", 4325), rep("sun_103", 4325), rep("103_100e", 4325), rep("103_101w", 4325), rep("100e_101w", 4325), rep("sun_100e", 5497), rep("sun_101w", 5497), rep("sun_103", 5497), rep("103_100e", 5497), rep("103_101w", 5497), rep("100e_101w", 5497))

ash_data_diffs$diff_cat <- as.factor(ash_data_diffs$diff_cat)

ash_data_diffs$rem_stat <- c(rep("before_removal", 25950), rep("after_removal", 32982))

ash_data_diffs$date_time[1:4325] <- c(ash_data3_before[,1])
ash_data_diffs$date_time[4326:8650] <- c(ash_data3_before[,1])
ash_data_diffs$date_time[8651:12975] <- c(ash_data3_before[,1])
ash_data_diffs$date_time[12976:17300] <- c(ash_data3_before[,1])
ash_data_diffs$date_time[17301:21625] <- c(ash_data3_before[,1])
ash_data_diffs$date_time[21626:25950] <- c(ash_data3_before[,1])

ash_data_diffs <- ash_data_diffs %>% mutate(date_time = ymd_hms(date_time))

ash_data_diffs$date_time[25951:31447] <- c(ash_data3_after[,1])
ash_data_diffs$date_time[31448:36944] <- (ash_data3_after[,1])
ash_data_diffs$date_time[36945:42441] <- (ash_data3_after[,1])
ash_data_diffs$date_time[42442:47938] <- (ash_data3_after[,1])
ash_data_diffs$date_time[47939:53435] <- (ash_data3_after[,1])
ash_data_diffs$date_time[53436:58932] <- (ash_data3_after[,1])


ash_data_diffs$diff_tempF[1:4325] <- c(ash_data3_before[,7])
ash_data_diffs$diff_tempF[4326:8650] <- c(ash_data3_before[,8])
ash_data_diffs$diff_tempF[8651:12975] <- c(ash_data3_before[,9])
ash_data_diffs$diff_tempF[12976:17300] <- c(ash_data3_before[,10])
ash_data_diffs$diff_tempF[17301:21625] <- c(ash_data3_before[,11])
ash_data_diffs$diff_tempF[21626:25950] <- c(ash_data3_before[,12])

ash_data_diffs$diff_tempF[25951:31447] <- c(ash_data3_after[,7])
ash_data_diffs$diff_tempF[31448:36944] <- c(ash_data3_after[,8])
ash_data_diffs$diff_tempF[36945:42441] <- c(ash_data3_after[,9])
ash_data_diffs$diff_tempF[42442:47938] <- c(ash_data3_after[,10])
ash_data_diffs$diff_tempF[47939:53435] <- c(ash_data3_after[,11])
ash_data_diffs$diff_tempF[53436:58932] <- c(ash_data3_after[,12])

ash_data_diffs$rem_stat <- factor(ash_data_diffs$rem_stat, levels = c('before_removal','after_removal'))

ggplot(data = ash_data_diffs) +
  geom_density(mapping = aes(x = diff_tempF, y = stat(density), color = diff_cat), position = "identity", bins = 36, alpha = 0.6) +
  facet_wrap(facets = ~ rem_stat) 

ggplot(data = ash_data_diffs) +
  geom_density(mapping = aes(x = diff_tempF, y = stat(density), color = diff_cat), position = "identity", bins = 36, alpha = 0.6) +
  xlim(-3, 3) +
  facet_wrap(facets = ~ rem_stat) 
```
               

*Step 2.2:* Run overall value t-tests between the hourly averaged differences in temperature before ash removal and after 
```{r t-test, echo = FALSE}
ash_data3_before[4326:5497, ] <- NA

t_sun_100e <- t.test(ash_data3_after$diff_sun_100e, 
                    ash_data3_before$diff_sun_100e, paired = TRUE)

psun100e <- t_sun_100e$p.value

paste("p-val for sun vs 100e is ", psun100e)

#########

t_sun_101w <- t.test(ash_data3_after$diff_sun_101w, 
                    ash_data3_before$diff_sun_101w, paired = TRUE)

psun101w <- t_sun_101w$p.value

paste("p-val for sun vs 101w is ", psun101w)

#########

t_sun_103 <- t.test(ash_data3_after$diff_sun_103, 
                    ash_data3_before$diff_sun_103, paired = TRUE)

psun103 <- t_sun_103$p.value

paste("p-val for sun vs 103 is ", psun103)

#########

t_103_100e <- t.test(ash_data3_after$diff_103_100e, 
                    ash_data3_before$diff_103_100e, paired = TRUE)

p103100e <- t_103_100e$p.value

paste("p-val for 103 vs 100e is ", p103100e)

#########

t_103_101w <- t.test(ash_data3_after$diff_103_101w, 
                    ash_data3_before$diff_103_101w, paired = TRUE)

p103101w <- t_103_101w$p.value

paste("p-val for 103 vs 101w is ", p103101w)

#########

t_100e_101w <- t.test(ash_data3_after$diff_100e_101w, 
                    ash_data3_before$diff_100e_101w, paired = TRUE)

p100e101w <- t_100e_101w$p.value

paste("p-val for 100e vs 101w is ", p100e101w)
```


*Step 3:* Make a barplot or boxplot to show the average difference and st error in mean temperatures in the before and after treatments.
```{r bar-plot-avg-diffs-and-ST-DEVIATION, echo=FALSE}

melted <- melt(ash_data_diffs, id.vars = c("diff_cat", "rem_stat", "date_time"))

mean_diffs <- ddply(melted, c("diff_cat", "rem_stat"), summarise, mean = mean(value), sd = sd(value))

mean_diffs$diff_cat <- as.factor(mean_diffs$diff_cat)

ggplot(data = mean_diffs, aes(x = diff_cat, y = mean, color = diff_cat, fill = diff_cat)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), width = .2) +
  facet_wrap(facets = ~rem_stat)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```


```{r bar-plot-avg-diffs-and-ST-ERROR, echo=FALSE}

melted <- melt(ash_data_diffs, id.vars = c("diff_cat", "rem_stat", "date_time"))

### ??? can I export this table
mean_diffs <- ddply(melted, c("diff_cat", "rem_stat"), summarise, mean = mean(value), sd = sd(value))

#create unique diff_cat names in each row

melted1a <- melted %>% 
  filter(diff_cat == "sun_103" & rem_stat == "before_removal") %>% 
  mutate(diff_cat = "sun_103_before")
melted1b <- melted %>% 
  filter(diff_cat == "sun_103" & rem_stat == "after_removal") %>% 
  mutate(diff_cat = "sun_103_after")
melted2a <- melted %>% 
  filter(diff_cat == "sun_101w" & rem_stat == "before_removal") %>% 
  mutate(diff_cat = "sun_101w_before")
melted2b <- melted %>% 
  filter(diff_cat == "sun_101w" & rem_stat == "after_removal") %>% 
  mutate(diff_cat = "sun_101w_after")
melted3a <- melted %>% 
  filter(diff_cat == "sun_100e" & rem_stat == "before_removal") %>% 
  mutate(diff_cat = "sun_100e_before")
melted3b <- melted %>% 
  filter(diff_cat == "sun_100e" & rem_stat == "after_removal") %>% 
  mutate(diff_cat = "sun_100e_after")
melted4a <- melted %>% 
  filter(diff_cat == "103_101w" & rem_stat == "before_removal") %>% 
  mutate(diff_cat = "103_101w_before")
melted4b <- melted %>% 
  filter(diff_cat == "103_101w" & rem_stat == "after_removal") %>% 
  mutate(diff_cat = "103_101w_after")
melted5a <- melted %>% 
  filter(diff_cat == "103_100e" & rem_stat == "before_removal") %>% 
  mutate(diff_cat = "103_100e_before")
melted5b <- melted %>% 
  filter(diff_cat == "103_100e" & rem_stat == "after_removal") %>% 
  mutate(diff_cat = "103_100e_after")
melted6a <- melted %>% 
  filter(diff_cat == "100e_101w" & rem_stat == "before_removal") %>% 
  mutate(diff_cat = "100e-101w_before")
melted6b <- melted %>% 
  filter(diff_cat == "100e_101w" & rem_stat == "after_removal") %>% 
  mutate(diff_cat = "100e-101w_after")

# bind together back into 1 dataframe
melted_renamed <- rbind(melted1a, melted1b, melted2a, melted2b, melted3a, melted3b, melted4a, melted4b, melted5a, melted5b, melted6a, melted6b)

# calculate summary statistics
melted_summary <- melted_renamed %>% 
  group_by(diff_cat) %>% 
  summarise(mean_temp_diff = mean(value), # calc mean for each diff category
            sd_temp_dif = sd(value), # calc st. deviation
            n_logs = n(), # calc sample size
            SE_temp_diff = sd(value)/sqrt(n())) %>%  # calc st error of each group
  mutate(rem_stat = rep("before_removal"))
# fina a better way to change these factor values, manuel for now
melted_summary$rem_stat[1] = "after_removal"
melted_summary$rem_stat[3] = "after_removal"
melted_summary$rem_stat[5] = "after_removal"
melted_summary$rem_stat[7] = "after_removal"
melted_summary$rem_stat[9] = "after_removal"
melted_summary$rem_stat[11] = "after_removal"


# make a bar graph with st error bars
temp_diff_barplot <- ggplot(melted_summary, aes(diff_cat, mean_temp_diff)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_temp_diff-SE_temp_diff, ymax = mean_temp_diff + SE_temp_diff), width = 0.2) 


# rename the diff catogries to have just the diff groups, not the removal status
melted_summary2 <- melted_summary %>% 
  mutate( diff_cat = c("100e-101w", "100e-101w", "103_100e", "103_100e", "103_101w", "103_101w", "sun_100e", "sun_100e", "sun_101w", "sun_101w", "sun_103", "sun_103"))
# return to factor
melted_summary2$diff_cat <- as.factor(melted_summary2$diff_cat)

# change the order of the removal status factor
melted_summary2$rem_stat <- factor(melted_summary2$rem_stat, levels = c("before_removal", "after_removal"))


# color code and facet the graph

p1 <- ggplot(data = melted_summary2, aes(x = diff_cat, y = mean_temp_diff, color = diff_cat, fill = diff_cat)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(ymin = mean_temp_diff-SE_temp_diff, ymax = mean_temp_diff+SE_temp_diff), width = .2) +
  facet_wrap(facets = ~rem_stat)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ylab("mean difference in air temperature") +  # yaxis labels 
  xlab("temperature sensor location pairs")  # xaxis label




p2 <- ggplot(data = melted_summary2, aes(x = diff_cat, y = mean_temp_diff,
                                        color = factor(diff_cat),
                                        fill = factor(diff_cat))) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(ymin = mean_temp_diff-SE_temp_diff, ymax = mean_temp_diff+SE_temp_diff), width = .2) +
  facet_wrap(facets = ~rem_stat)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ylab("mean difference in air temperature (F)") +  # yaxis labels 
  xlab("temperature sensor location pairs")  # xaxis label

p2 + 
  scale_color_manual(name = "paired differences",
                     labels = c("100e-101w (treatment-treatment)",
                                "103-100e (control-treatment)",
                                "103-101w (control-treatment)",
                                "sun-100e (control-treatment)",
                                "sun-101w (control-treatment)",
                                "sun-103 (control-control)"),
                     values = c("100e-101w" = "burlywood4",
                                "103_100e" = "green4",
                                "103_101w" = "seagreen4",
                                "sun_100e" = "coral4",
                                "sun_101w" = "salmon2",
                                "sun_103" = "grey49")) +
  scale_fill_manual(name = "paired differences",
                     labels = c("100e-101w (treatment-treatment)",
                                "103-100e (control-treatment)",
                                "103-101w (control-treatment)",
                                "sun-100e (control-treatment)",
                                "sun-101w (control-treatment)",
                                "sun-103 (control-control)"),
                     values = c("100e-101w" = "burlywood4",
                                "103_100e" = "green4",
                                "103_101w" = "seagreen4",
                                "sun_100e" = "coral4",
                                "sun_101w" = "salmon2",
                                "sun_103" = "grey49"))

```

**Step 2.3 ANOVA **
```{r - anova}
library(lme4)
# run a linear model
lm1 <- lm(value ~ diff_cat, data = melted_renamed) 
summary(lm1)

# run an anova
aov1 <- aov(value ~ diff_cat, data = melted_renamed)
summary(aov1)


# Tukey's test to look at pairwise comparisons
tukey.test <- TukeyHSD(aov1)
tukey.test


```



```{r merge-difference-to-single-graph}
# add before after difference
melted_summary3 <- melted_summary %>%
  mutate(b_a_diff = c("0.2695", "0.2695", "-0.4724", "-0.4724", "-0.2029", "-0.2029", "-1.0228", "-1.0228", "-0.7532", "-0.7532", "-0.5504", "-0.5504"))

# make a bar graph with st error bars
ggplot(melted_summary3, aes(diff_cat, b_a_diff)) +
  geom_col() 
  geom_errorbar(aes(ymin = b_a_diff-SE_temp_diff, ymax = mean_temp_diff + SE_temp_diff), width = 0.2) 


```


```{r}

# make a bar graph with st error bars
temp_diff_barplot <- ggplot(melted_summary, aes(diff_cat, mean_temp_diff)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_temp_diff-SE_temp_diff, ymax = mean_temp_diff + SE_temp_diff), width = 0.2)

```



```{r box-plot-avg-diffs, echo=FALSE}

melted <- melt(ash_data_diffs, id.vars = c("diff_cat", "rem_stat", "date_time"))

mean_diffs <- ddply(melted, c("diff_cat", "rem_stat"), summarise, mean = mean(value), se = sd(value))

mean_diffs$diff_cat <- as.factor(mean_diffs$diff_cat)
mean_diffs$sensor_differences <- as.factor(mean_diffs$diff_cat)

ggplot(data = mean_diffs, aes(x = sensor_differences, y = mean, color = sensor_differences, fill = sensor_differences)) +
  geom_boxplot(stat = 'boxplot') +
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width = .2) +
  geom_hline(yintercept = 0) +
  facet_wrap(facets = ~rem_stat)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

*Step 4:* Make a line plot of the differences in temperature in the before and after treatments in relation to days
    - may need to change "date_time" to day 1, 2, 3 so able to fully overlay values
```{r line-plot-diffs, echo = FALSE}

ash_data_diffs %>%
  group_by(rem_stat) %>%
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = diff_tempF, color = diff_cat), size =0.3) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) + 
  geom_vline(xintercept = ymd_hms("2018-09-20 18:22:59")) 
```

```{r overlay-line-plots-2, echo = FALSE}
ash_data_diffs %>% 
  group_by(rem_stat) %>%
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = diff_tempF, color = diff_cat)) +
  facet_wrap(facets = ~ rem_stat) 
```

