---
title: "Preliminary Data from Boulder Ash Removal Microclimate Study"
author: "Deidre Jaeger"
date: "11/9/2018"
output: pdf_document
---

??? add introduction

**Research question:** How is microclimate impacted by ash (*Fraxinus sp.*) removal? 

**Hypothesis:** Air temperature will increase at the location of the removed ash trees (#100e and 101w) due to increased solar radiation and decreasing shade and evapotranspiration. 

**Methods:** Air temperature was measured in four locations along Boulder creek (Photo 1.) from September 17-24. Removal of the adjacent ash trees #100e and 101w occured on September 20th between 8am - 11am.

??? add diagram of study design


```{r global_options, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE, results= FALSE)

```


```{r load-libraries, results= FALSE}
# use install.packages("insert-package-name-here") if you get an error message that says "there is no package called ....x...."

library(dplyr) # for mutate() functionality
library(lubridate) # for as.POSIX function
library(ggplot2)
library(gridExtra) # for side by side plots
library(png) # libraries for images
library(grid) # libraries for images
options(stringsAsFactors = FALSE) # so that dates read in as character strings not factors (factors are troublesome for converting to dates)
```
 
```{r photo1, echo=FALSE, fig.cap= "Photo 1. Aerial view of the study area, North of Boulder Creek, Boulder, Colorado."}
img1 <- readPNG("/Users/deidrejaeger/Google Drive/UrbanEcologyResearch/Projects/BoulderAsh/ash-microclimate-study/images/aerial_map_study_area.png")
 grid.raster(img1)

```
 
 
```{r step1-import-treatment-data}

# import the data from the treatment and control sensors

# open the first file of the treatment-tree 100e, from the "before" removal
treatment_100e_before <- read.csv("data/SN 20439121 2018-09-20 00_06_33 -0600.csv", # define the file name
                          skip = 1, # skip reading the first row
                          header = TRUE, # define the second row as the header
                          col.names = c("date_time", "tempF", "sensor_sn","sensor_name", "ash_removal", "blank1")) # rename the columns
treatment_100e_before$blank1 <- NULL # Remove the blank columns
treatment_100e_before$sensor_sn <- rep(20439121) # add the sensor serial number
treatment_100e_before$sensor_name <- rep("treatment_tree_100e") # fill column with the sensor name
treatment_100e_before$ash_removal <- rep("before_removal") # fill the column with the ash removal status


# check the first 6 rows to see if the dataframe looks as we expect
head(treatment_100e_before)

# check the structure of how it is loaded
str(treatment_100e_before) # we will eventually want date to be formated as date, but for now having number for temp and character for sensor name look good


# open the second file of the treatment-tree 100e, from the "after" removal
# open the first file of the treatment-tree 100e, from the "before" removal
treatment_100e_after <- read.csv("data/SN 20439121 2018-09-26 10_27_52 -0600.csv", # define the file name
                          skip = 1, # skip reading the first row
                          header = TRUE, # define the second row as the header
                          col.names = c("date_time", "tempF", "sensor_sn","sensor_name", "ash_removal", "blank1")) # rename the columns
treatment_100e_after$blank1 <- NULL # Remove the blank columns
treatment_100e_after$sensor_sn <- rep(20439121) # add the sensor serial number
treatment_100e_after$sensor_name <- rep("treatment_tree_100e") # fill column with the sensor name
treatment_100e_after$ash_removal <- rep("after_removal") # fill the column with the ash removal status

# check the first 6 rows to see if the dataframe looks as we expect
head(treatment_100e_after)

# combine the before and after treatment into 1 dateframe
treatment_100e_clean <- rbind(treatment_100e_before, treatment_100e_after)

# check the structure of how it is loaded
str(treatment_100e_clean) # date is a character, we want to code it as a date

# format 
treatment_100e_clean <- treatment_100e_clean %>% 
  mutate(date_time = ymd_hms(date_time)) # format date in R-friendly format

str(treatment_100e_clean) # check that class structure changes to "POSIXct"


#### tree 101w
# import the data from the treatment and control sensors

# open the first file of the treatment-tree 100e, from the "before" removal
treatment_101w_before <- read.csv("data/SN 20439119 2018-09-20 00_07_03 -0600.csv", # define the file name
                          skip = 1, # skip reading the first row
                          header = TRUE, # define the second row as the header
                          col.names = c("date_time", "tempF", "sensor_sn","sensor_name", "ash_removal", "blank1")) # rename the columns
treatment_101w_before$blank1 <- NULL # Remove the blank columns
treatment_101w_before$sensor_sn <- rep(20439119) # add the sensor serial number
treatment_101w_before$sensor_name <- rep("treatment_tree_101w") # fill column with the sensor name
treatment_101w_before$ash_removal <- rep("before_removal") # fill the column with the ash removal status


# check the first 6 rows to see if the dataframe looks as we expect
head(treatment_101w_before)

# check the structure of how it is loaded
str(treatment_101w_before) # we will eventually want date to be formated as date, but for now having number for temp and character for sensor name look good


# open the second file of the treatment-tree 100e, from the "after" removal
# open the first file of the treatment-tree 100e, from the "before" removal
treatment_101w_after <- read.csv("data/SN 20439119 2018-09-26 10_22_57 -0600.csv", # define the file name
                          skip = 1, # skip reading the first row
                          header = TRUE, # define the second row as the header
                          col.names = c("date_time", "tempF", "sensor_sn","sensor_name", "ash_removal", "blank1")) # rename the columns
treatment_101w_after$blank1 <- NULL # Remove the blank columns
treatment_101w_after$sensor_sn <- rep(20439119) # add the sensor serial number
treatment_101w_after$sensor_name <- rep("treatment_tree_101w") # fill column with the sensor name
treatment_101w_after$ash_removal <- rep("after_removal") # fill the column with the ash removal status

# check the first 6 rows to see if the dataframe looks as we expect
head(treatment_101w_after)

# combine the before and after treatment into 1 dateframe
treatment_101w_clean <- rbind(treatment_101w_before, treatment_101w_after)

# check the structure of how it is loaded
str(treatment_101w_clean) # date is a character, we want to code it as a date

# format 
treatment_101w_clean <- treatment_101w_clean %>% 
  mutate(date_time = ymd_hms(date_time)) # format date in R-friendly format

str(treatment_101w_clean) # check that class structure changes to "POSIXct"

```


```{r step2-import-control-data-files}
# open the first file of the treatment-tree 100e, from the "before" removal
control_103_data <- read.csv("data/SN 20397886 2018-09-26 10_27_32 -0600.csv", # define the file name
                          skip = 1, # skip reading the first row
                          header = TRUE, # define the second row as the header
                          col.names = c("date_time", "tempF", "sensor_sn","sensor_name", "ash_removal", "blank1")) # rename the columns

control_103_data$blank1 <- NULL # Remove the blank columns
control_103_data$sensor_sn <- rep(20397886) # add the sensor serial number
control_103_data$sensor_name <- rep("control_tree_103") # fill column with the sensor name

# format date/time
control_103_data <- control_103_data %>% 
  mutate(date_time = ymd_hms(date_time)) # format date in R-friendly format

# assign "before" ash_removal categories
control_103_data_before <- control_103_data %>% 
  filter(date_time <= ymd_hms("2018-09-20 00:03:59")) %>% 
  mutate(ash_removal = "before_removal")

# assign "during" ash_removal categories ( Deidre needs to fix this still!)
control_103_data_during <- control_103_data %>% 
  filter((date_time >= ymd_hms("2018-09-20 00:04:00")) & (date_time <= ymd_hms("2018-09-20 12:20:59"))) %>% 
  mutate(ash_removal = "during_removal")

# assign "after" ash_removal categories
control_103_data_after <- control_103_data %>% 
  filter(date_time >= ymd_hms("2018-09-20 12:21:00")) %>% 
  mutate(ash_removal = "after_removal")

# zip back up into 1 dataframe
control_103_data_clean <- rbind(control_103_data_before, control_103_data_during, control_103_data_after)

# check that all rows were assigned an ash removal status
unique(control_103_data_clean$ash_removal)# Yay! There are no NA's meaning we captured all the rows
nrow(control_103_data) == nrow(control_103_data_clean) # and we didn't lose any rows!


##### sun control
# open the first file of the treatment-tree 100e, from the "before" removal
control_sun_data <- read.csv("data/SN 20439118 2018-09-26 10_29_52 -0600.csv", # define the file name
                          skip = 1, # skip reading the first row
                          header = TRUE, # define the second row as the header
                          col.names = c("date_time", "tempF", "sensor_sn","sensor_name", "ash_removal", "blank1")) # rename the columns

control_sun_data$blank1 <- NULL # Remove the blank columns
control_sun_data$sensor_sn <- rep(20439118) # add the sensor serial number
control_sun_data$sensor_name <- rep("control_sun") # fill column with the sensor name

# format date/time
control_sun_data <- control_sun_data %>% 
  mutate(date_time = ymd_hms(date_time)) # format date in R-friendly format

# assign "before" ash_removal categories
control_sun_data_before <- control_sun_data %>% 
  filter(date_time <= ymd_hms("2018-09-20 00:03:59")) %>% 
  mutate(ash_removal = "before_removal")

# assign "during" ash_removal categories ( Deidre needs to fix this still!)
control_sun_data_during <- control_sun_data %>% 
  filter((date_time >= ymd_hms("2018-09-20 00:04:00")) & (date_time <= ymd_hms("2018-09-20 12:20:59"))) %>% 
  mutate(ash_removal = "during_removal")

# assign "after" ash_removal categories
control_sun_data_after <- control_sun_data %>% 
  filter(date_time >= ymd_hms("2018-09-20 12:21:00")) %>% 
  mutate(ash_removal = "after_removal")

# zip back up into 1 dataframe
control_sun_data_clean <- rbind(control_sun_data_before, control_sun_data_during, control_sun_data_after)

# check that all rows were assigned an ash removal status
unique(control_sun_data_clean$ash_removal)# Yay! There are no NA's meaning we captured all the rows
nrow(control_sun_data) == nrow(control_sun_data_clean) # and we didn't lose any rows!

```

```{r step3-bind-into-1-csv}

# aggregate the before and after dataframes for all sensor data into 1 dataframe (best practice for tidy analyses)

# for example:
all_data <- rbind(treatment_100e_clean, treatment_101w_clean, control_103_data_clean, control_sun_data_clean)

```
 
```{r step4-set-factors}

# we want the sensor serial number, sensor name, and the ash removal status columns to be categorical factors, rather than text characters

# for example:
all_data <- all_data %>% 
  mutate(sensor_sn = as.factor(sensor_sn), sensor_name = as.factor(sensor_name), ash_removal = as.factor(ash_removal))

str(all_data) # check that it worked
```

```{r step5-remove-temp-NAs}

# check for NAs in dataset
all_data %>% 
  filter(is.na(tempF)) # fifteen rows don't have a temp value

# remove the rows that have NA temperature values with (! is a boolean operator and means "is not equal to"
all_data <- all_data %>% 
  filter(!is.na(tempF))

```

```{r step6-export-csv}
# export this new dataframe as "ash-microclimate-data2018" so we don't have to rebuild it each time we want to work with it, or we can send it someone over email

# for example: 
all_data %>% 
   write.csv(file="data/ash-microclimate-data2018-prelim.csv") # using pipes, write a new csv file saved at the defined filepath
  
```

  
```{r load-data, echo=FALSE}

# read csv file
ash_data <- read.csv("data/ash-microclimate-data2018-prelim.csv", header = T) 

# check structure and make sure every column and row we expect to be there is present
head(ash_data)
unique(ash_data$sensor_name)
str(ash_data) # darn, we dont have the factors and date formatted anymore

# reset date and factors
ash_data <- ash_data %>% 
  mutate(date_time = ymd_hms(date_time), sensor_sn = as.factor(sensor_sn), sensor_name = as.factor(sensor_name), ash_removal = as.factor(ash_removal))

# recheck structure to make sure date=POSIXct, tempF = num, and everything else is a factor- except we don't care about the "X" factor
str(ash_data) 

```

```{r view-histograms-by-sensor, results = TRUE, fig.cap = "Figure 1. Temperature value frequency distributions from each sensor during each treatment phase"}

ash_data %>% # define date
  ggplot() +
  geom_histogram(mapping = aes(x= tempF,
                               color = sensor_name), bins = 40) + # define x axis
  facet_grid(facets = sensor_name ~ ash_removal) # define which variables to panel vertical and horizontal graphs with

```
Figure 1. 

```{r view-time-series-lineplot, results = TRUE, fig.cap = "Figure 2. Time series plot showing control and treatment sensors over the study duration, Sept 17- Sept 24, 2018. Black vertical lines indicate time of tree removal, treatment tree sensors were not logging during this time removal time"}

ash_data %>%
  filter(ash_removal == "before_removal" | ash_removal == "after_removal") %>% 
  group_by(ash_removal) %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
                           color = ash_removal)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59")) +
  facet_wrap( facets = ~ sensor_name)

```

```{r control-only-plots,  results = TRUE, fig.cap = "Time series plot showing control sensors over the study duration, Sept 17- Sept 24, 2018. Black vertical lines indicate time of tree removal, treatment tree sensors were not logging during this time removal time"}
ash_data %>% 
  filter(sensor_name == "control_sun" | 
           sensor_name == "control_tree_103") %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
            color = sensor_name,
            group = sensor_name,
            linetype = sensor_name)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59"))
```

```{r treatment-only-plots, results = TRUE,  fig.cap = "Figure 3. Time series plot showing individual treatment trees 100 and 101 over the study duration, Sept 17- Sept 24, 2018. Black vertical lines indicate time of tree removal, treatment tree sensors were not logging during this time removal time"}

# make the 100e treatment tree
ash_data %>% 
  filter(sensor_name == "treatment_tree_100e" | 
           sensor_name == "treatment_tree_101w") %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
            color = sensor_name,
            group = sensor_name,
            linetype = sensor_name)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59"))


```

```{r treatment-separated-plots,  results = TRUE, fig.cap = "Figure 4. Time series plot showing individual treatment tree and control sensors over the study duration, Sept 17- Sept 24, 2018. Black vertical lines indicate time of tree removal, treatment tree sensors were not logging during this time removal time"}

# make the 100e treatment tree
data100e <- ash_data %>% 
  filter(sensor_name == "treatment_tree_100e" | 
           sensor_name == "control_sun" | 
           sensor_name == "control_tree_103") %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
            color = sensor_name,
            group = sensor_name,
            linetype = sensor_name)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59"))

# make the 101w treatment tree
data101w <- ash_data %>% 
  filter(sensor_name == "treatment_tree_101w" | 
           sensor_name == "control_sun" | 
           sensor_name == "control_tree_103") %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
            color = sensor_name,
            group = sensor_name,
            linetype = sensor_name)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59"))

grid.arrange(data100e, data101w)
```

```{r view-grouped-overay,  results = TRUE, fig.cap = "Figure 5. Time series plot of all 4 sensors over the study duration, Sept 17- Sept 24, 2018. Black vertical lines indicate time of tree removal, treatment tree sensors were not logging during this time removal time"}

ash_data %>% 
  group_by(sensor_name) %>% 
  ggplot() +
  geom_line(mapping = aes(x = date_time, y = tempF,
            color = sensor_name,
            group = sensor_name,
            linetype = sensor_name)) +
  geom_vline(xintercept = ymd_hms("2018-09-20 00:04:00")) +
  geom_vline(xintercept= ymd_hms("2018-09-20 12:20:59")) 
  

```

```{r overlapping-histograms, echo = FALSE}

ash_data %>% 
  group_by(sensor_name) %>% 
ggplot(data = ash_data) +
  geom_histogram(mapping = aes(x = tempF, y = stat(density), fill = sensor_name), position = "identity", bins = 36, alpha = 0.6) +
  facet_wrap(facets = ~ sensor_name)
```

Next Steps:

- could standardize all the temps by subtracting from another locaiton weather station. 

```{r find-max-values}
ash_data %>%
  filter(ash_removal == "before_removal") %>% 
           group_by(sensor_name) %>% 
           filter(tempF == max("tempF")) 
     
# Look at the max values for the study duration
# before removal by sensor
ash_data %>%
  filter(ash_removal == "before_removal") %>% 
           group_by(sensor_name) %>% 
           arrange( desc(tempF)) %>%
            slice(1)

# top value timeframe comparison
max_before <- ash_data %>%
  filter(date_time == "2018-09-17 6:02:59") 


# after removal by sensor
ash_data %>%
  filter(ash_removal == "after_removal") %>% 
           group_by(sensor_name) %>% 
           arrange( desc(tempF)) %>%
            slice(1) 

# top value timeframe comparison
 max_after <- ash_data %>% 
        filter(date_time == "2018-09-23 7:22:59" | date_time == "2018-09-23 7:22:21" )  # bc the sensors weren't logging every minute at the same second mark


max_vals <- rbind(max_before, max_after)
max_vals$alph_stat <- factor(max_vals$ash_removal, levels= c("before_removal", "after_removal", "during_removal"))



```

```{r barplot-max-values}
max_vals %>%
  group_by(ash_removal) %>% 
  ggplot() +
  geom_bar(mapping = aes(x = sensor_name, y = tempF, fill= sensor_name), 
           stat= "identity") +
           coord_cartesian(ylim=c(82,92)) +
  geom_text(aes(label = tempF, x = sensor_name, y = tempF, fill= sensor_name)) +
  facet_grid(alph_stat ~ .)
```


```{r find-min-values}

     # Look at the max values for the study duration
# before removal by sensor
ash_data %>%
  filter(ash_removal == "before_removal") %>% 
           group_by(sensor_name) %>% 
           arrange(tempF) %>%
            slice(1)

# top value timeframe comparison
min_before <- ash_data %>%
  filter(date_time == "2018-09-19 1:25:59") 


# after removal by sensor
ash_data %>%
  filter(ash_removal == "after_removal") %>% 
           group_by(sensor_name) %>% 
           arrange(tempF) %>%
            slice(1) 

# top value timeframe comparison
 min_after <- ash_data %>% 
        filter(date_time == "2018-09-22 1:12:59" | date_time == "2018-09-22 1:12:21" )  # bc the sensors weren't logging every minute at the same second mark


min_vals <- rbind(min_before, min_after)
min_vals$alph_stat <- factor(min_vals$ash_removal, levels= c("before_removal", "after_removal", "during_removal"))



```

```{r barplot-min-values}
min_vals %>%
  group_by(ash_removal) %>% 
  ggplot() +
  geom_bar(mapping = aes(x = sensor_name, y = tempF, fill= sensor_name), 
           stat= "identity") +
           coord_cartesian(ylim=c(42,58)) +
  geom_text(aes(label = tempF, x = sensor_name, y = tempF, fill= sensor_name)) +
  facet_grid(alph_stat ~ .)
```

```{r}

```

